---
title: "Case Study: How Does a Bike-Share Navigate Speedy Success?"
author: "Max Dotson"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

### Scenario

You are a junior data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company's future success depends on maximizing the number of annual memberships. Therefore, your team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations.

### About the company

In 2016, Cyclistic launched a successful bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that are geotracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to any other station in the system anytime.

### Business Task

Design marketing strategies aimed at converting casual riders into annual members.

### Prepare

Cyclistic's historical trip data (September 2021 - September 2022) has been utilized for business findings, and it is organized in 13 .CSV files. The data was made available by Motivate International Inc. under this [license](https://divvy-tripdata.s3.amazonaws.com/index.html).)

### Process

-   R programming is utilized for the processing and analysis of data due to its capability to handle large data sets, and visualization tools.

```{r}
# Installation and loading of necessary packages
install.packages("tidyverse")
install.packages("lubridate")
install.packages("ggplot2")

library(tidyverse)
library(lubridate)
library(ggplot2)
```

```{r}
#=====================
# STEP 1: COLLECT DATA
#=====================

sep_21<-read_csv("202109-divvy-tripdata.csv")
oct_21<-read_csv("202110-divvy-tripdata.csv")
nov_21<-read_csv("202111-divvy-tripdata.csv")
dec_21<-read_csv("202112-divvy-tripdata.csv")
jan_22<-read_csv("202201-divvy-tripdata.csv")
feb_22<-read_csv("202202-divvy-tripdata.csv")
mar_22<-read_csv("202203-divvy-tripdata.csv")
apr_22<-read_csv("202204-divvy-tripdata.csv")
may_22<-read_csv("202205-divvy-tripdata.csv")
jun_22<-read_csv("202206-divvy-tripdata.csv")
jul_22<-read_csv("202207-divvy-tripdata.csv")
aug_22<-read_csv("202208-divvy-tripdata.csv")
sep_22<-read_csv("202209-divvy-publictripdata.csv")
```

```{r}
#====================================================
# STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE
#====================================================

sep_21<- mutate(sep_21, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 
oct_21<- mutate(oct_21, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 
nov_21<- mutate(nov_21, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 
dec_21<- mutate(dec_21, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 
jan_22<- mutate(jan_22, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 
feb_22<- mutate(feb_22, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 
mar_22<- mutate(mar_22, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 
apr_22<- mutate(apr_22, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 
may_22<- mutate(may_22, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 
jun_22<- mutate(jun_22, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 
jul_22<- mutate(jul_22, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 
aug_22<- mutate(aug_22, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 
sep_22<- mutate(sep_22, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 

# All months joined into one data frame
all_trips<- bind_rows(sep_21, oct_21, nov_21, dec_21, jan_22, feb_22, mar_22, apr_22, may_22, jun_22, jul_22, aug_22, sep_22)
```

##### Step 3: Clean up data and prepare for analysis

-   Removal of columns listing latitudes and longitudes

```{r}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng))
```

-   Adding columns that list the date, month, day, and year of each ride

```{r}
all_trips$date <- as.Date(all_trips$started_at) 
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

-   Adding a "ride_length" calculation to all_trips (in seconds)

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

-   Converting "ride_length" from Factor to numeric so calculations can be run on the data

```{r}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
```

-   Removing "bad" data
-   The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

### Analyze/Share

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
```

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)	
```

##### Key Findings:

-   Casual riders spend a greater amount of time per ride than those who who are members.
-   Saturday and Sunday are the most popular days for casual riders, and also see an increase in average duration per ride.
-   Members take a greater amount of trips during the weekdays than they do the weekends, which may indicate that a majority of memberships are for work purposes.

```{r echo=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  labs(title = "Cyclistic Bike-Share Rides: September 2021 - September 2022")+
  theme(plot.title = element_text(size = 11))+ 
  geom_col(position = "dodge")
```

```{r echo=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  labs(title = "Average Cyclistic Ride-Share Duration: September 2021 - September 2022", subtitle = "Duration measured in seconds")+
  theme(plot.title = element_text(size = 11))+
  theme(plot.subtitle = element_text(size = 9))+ 
  geom_col(position = "dodge")
```

### Act

##### Top 3 recommendations:

-   Implement a weekend focused membership tier that offers discounts to casual riders who switch to members
-   Allow money off a membership each month for those who average approx. 1700 (28.3 minutes) seconds of travel per day in a week.
-   Offer a chance for casual riders to ride new bikes on weekends, if they switch to a yearly, monthly, or weekend focused membership.
